---
title: "The adult mosquito model RMG"
subtitle: "A Behavioral State Model"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document: 
    theme: paper 
vignette: >
  %\VignetteIndexEntry{The adult mosquito model RMG}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, echo=FALSE, warning=FALSE, messages=FALSE, purl=F}
suppressMessages(library(knitr))
library(exDE) 
library(mobwork) 
library(deSolve) 
```

```{r, echo=F, eval=F}
devtools::load_all()
devtools::load_all("../../exDE")
```

Most of the models developed to model malaria parasite infections in mosquitoes look at the *infection states:* uninfected; infected; or infected and infectious. A few models have been developed that track also track *parity.* A different class of models tracks the behavioral / physiological state of mosquitoes, so we call them *behavioral state models.* In this model, mosquitoes are either *blood feeding* or *egg laying,* following Le Menach, *et al.* (2005)^[Menach AL, *et al.* The unexpected importance of mosquito oviposition behaviour for malaria: non-productive larval habitats can be sources for malaria transmission. Malar J 4, 23 (2005). https://doi.org/10.1186/1475-2875-4-23]. The model also has *infection states.* 

## The Model

This is a patch-based model with $p$ patches, and we assume that all the parameters, variables, and terms are of length $p$ except for $\Omega_b$ and $\Omega_q$, which are $p \times p$ matrices. 

### Variables

+ $U_b$ - uninfected, blood feeding mosquitoes

+ $U_q$ - uninfected, egg laying mosquitoes

+ $Y_b$ - infected but not infective, blood feeding mosquitoes

+ $Y_q$ - infected but not infective, egg laying mosquitoes

+ $Z_b$ - infective, blood feeding mosquitoes

+ $Z_q$ - infective, egg laying mosquitoes

### Terms 

+ $\Lambda$ -  the emergence rate of adult mosquitoes from aquatic habitats in each patch

+ $\kappa$ - the net infectiousness of humans, the probability a mosquito becomes infected after blood feeding on a human  

### Parameters


+ $f$ - the blood feeding rate 

+ $q$ - the human blood feeding fraction 

+ $\nu$ - the egg laying rate 

+ $g$ - the mosquito death rate, per mosquito

+ $\varphi$ - the rate that infected mosquitoes become infective, the inverse of the EIP

+ $\sigma_b$ - the patch emigration rate for blood-feeding mosquitoes

+ $\sigma_q$ - the patch emigration rate for egg-laying mosquitoes 

+ ${\cal K}_b$ - the dispersal matrix for blood-feeding mosquitoes 

+ ${\cal K}_q$ - the dispersal matrix for egg-laying mosquitoes

+ $\Omega_b$ - the demographic matrix for blood feeding mosquitoes; letting $I$ denote the identity matrix, 
$$\Omega_b = \mbox{diag}\left(g\right) - \mbox{diag}\left(\sigma_b\right) \left(I - \cal K_b \right)$$

+ $\Omega_q$ - the demographic matrix for egg laying mosquitoes; letting $I$ denote the identity matrix,  
$$\mbox{diag} \left(g\right) - \mbox{diag} \left(\sigma_q\right) \left(I - \cal K_q \right)$$

### Equations 

$$
\begin{array}{rl}
\dfrac{dU_b}{dt} &= \Lambda  + \nu U_g - f U_b - \Omega_b \cdot U_b \\
\dfrac{dU_g}{dt} &= f (1- f q \kappa) U_b - \nu U_g  - \Omega_g \cdot U_g \\
\dfrac{dY_b}{dt} &= \nu Y_g + \phi Y_g - (f+\varphi) Y_g - \Omega_b \cdot Y_b \\
\dfrac{dY_g}{dt} &= f q \kappa U_b + f Y_b - (\nu + \varphi) Y_g - \Omega_g \cdot Y_g  \\
\dfrac{dZ_b}{dt} &= \varphi Y_b + \nu Z_g - f Z - \Omega_b \cdot Z_b \\
\dfrac{dZ_g}{dt} &= \varphi Y_g + f Z  - \nu Z  - \Omega_q \cdot Z_q 
\end{array}
$$

## Implementation

The `xde_setup()` utilities allow the user to pass a single version of the dispersal matrix $\cal K.$ During `xde_setup()`, `Omega_b` and `Omega_q` are identical.  


```{r}
model <- xde_setup(MYZname="RMG", Lname="trace", Xname = "trace", nPatches=3)
```

```{r}
model <- xde_solve(model)
```

```{r, fig.height=7.5, fig.width=5.5}
par(mfrow = c(2,1))
xde_plot_M(model)
xde_plot_YZ(model, add_axes = F)
xde_plot_YZ_fracs(model)
```



